Functional timeseries reveal no systemic shifts in vertebrate communities

### 2021-07-26 paper/manuscript.Rmd
Rewrite intro, add conceptual figure

---
title: "Functional timeseries reveal no systemic shifts in vertebrate communities"
author: 
  - name: "Kari E A Norman"
    affiliation: ucb
  - name: "Carl Boettiger"
    affiliation: ucb, ropensci
  - name: "Perry de Valpine"
    affiliation: ucb
address:
  - code: ucb
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
  - code: ropensci
    address: "The rOpenSci Project, University of California Berkeley, Berkeley CA 94720-3114, USA" 
abstract: |

@@

knitr::opts_chunk$set(cache=FALSE, message=FALSE, warning=FALSE)

@@

Functional diversity offers a potentially powerful alternative for detecting and describing community change by providing a mechanistic link between the organisms in a system and the processes they perform [@mcgill2006].
By describing the functional trait space rather than species, functional diversity metrics capture the disproportionate impact of losses or gains of functionally unique species.
Functional diversity metrics are therefore particularly well-suited for assessing the implications of community shifts in the full range of species change scenarios.

@@

While functional loss is frequently cited as one of the most pressing concerns of the anthropocene (CITATIONS), observed and projected functional diversity trends are less certain.
Forecasts of functional loss range from negligible [@gallagher2013] to dire [@petchey2002; @pimiento2020]. And while some observed trends show significant functional loss [@flynn2009] others document no loss even in some of the most heavily impacted communities [@matuoka2020]. Overall trends are therefore necessary... Clearly functional loss is not inevitable in scenarios of perturbation or even species loss [@daz2001].

<!-- trends in functional composition also have important implications for how we assesses potential changes in ecosystem stability, function, and resilience. Functional loss is often implicit in work linking biodiversity loss to impacts on ecosystem processes despite a wide range of documented relationships between functional and species change in natural communities [@biswas2011]. Where functional diversity is explicitly considered, there is increasing evidence it is a better predictor of ecosystem function than species-based metrics [@cadotte2011; @gagic2015], and that different facets of functional diversity play essential roles in maintaining ecosystem stability [@craven2018; @morin2014]. Tracking the relationship between species and functional changes across a wide variety of communities therefore provides an opportunity to assess not only the prevalence of concurrent species and functional loss, but also the evidence underpinning hypothesis between biodiversity loss and ecosystem processes. -->

<!-- paragraph on maintenence of functional structure through time -->

@@

```{r taxa-map, fig.cap="Map of timeseries locations with points color corresponding to taxa.", echo=FALSE}
knitr::include_graphics(here::here("figures/study_map.jpeg"))
```

@@

@@

@@

Our study represents the largest broad-scale multi-taxa assessment of functional change through time to date, giving a first look at overall trends in functional diversity and the range of changes happening in mammal, bird, and amphibian communities.
Surprisingly, we did not detect an overall trend in any of the calculated functional diversity metrics.
While nearly all of the studies included in our data experienced a significant increase or decrease in at least one of the metrics, trends were observed in equal numbers resulting in an overall trend that was not significantly different from zero.
As with previous species-based syntheses, we also found no overall trend in species diversity accompanied by increasing turnover through time [@dornelas2018], indicating that non-significant trends in functional metrics may be consistent with similar well-documented species derived trends.
We find no evidence of systemic functional richness loss or functional change.
A lack of trend for data broken down by realm, biome, and taxonomic group gives further evidence that directional functional change is absent from all systems observed in our dataset.

>>>>

---
title: "Functional timeseries reveal no systemic shifts in vertebrate communities"
author: 
  - name: "Kari E A Norman"
    affiliation: ucb
  - name: "Perry de Valpine"
    affiliation: ucb
  - name: "Carl Boettiger"
    affiliation: ucb, ropensci
address:
  - code: ucb
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
  - code: ropensci
    address: "The rOpenSci Project, University of California Berkeley, Berkeley CA 94720-3114, USA" 
abstract: |

@@

knitr::opts_chunk$set(cache=FALSE, message=FALSE, warning=FALSE, out.width = "\\textwidth")

@@

Functional diversity offers a potentially powerful alternative for detecting and describing community change by providing a mechanistic link between the organisms in a system and the processes they perform [@mcgill2006].
By describing the functional trait space rather than species, functional diversity metrics capture the disproportionate impact of losses or gains of functionally unique species.
Functional diversity metrics are therefore uniquely suited for assessing the implications of community shifts in the full range of species change scenarios.

@@

While functional loss is frequently cited as one of the most pressing concerns of the anthropocene [@cardinale2012; @dirzo2014; @young2016], it is not necessarily inevitable even in scenarios of species loss [@daz2001].
Forecasts of functional loss range from negligible [@gallagher2013] to dire [@petchey2002; @pimiento2020].
And while some observed trends show significant functional loss [@flynn2009] others document no loss even in some of the most heavily impacted communities [@edwards2013; @matuoka2020].
It is therefore critical to establish whether or not functional loss is the general rule for communities, and at what rate functional loss is occurring.

@@

```{r taxa-map, fig.cap="Map of timeseries locations with points color corresponding to taxa.", echo=FALSE}
knitr::include_graphics(here::here("figures/study_map_hist.jpeg"))
```

@@

```{r metric-fi, fig.cap="Example distributions of species and abundance for a single trait axis for high and low values of our measured metrics.", echo=FALSE, strip.white=TRUE}
knitr::include_graphics(here::here("figures/conceptual_fig.jpeg"))
```

@@

We assessed the impact of study-level characteristics on slope estimates...

@@

Our study represents the largest broad-scale multi-taxa assessment of functional change through time to date, giving a first look at overall trends in functional diversity and the range of changes happening in mammal, bird, and amphibian communities.
Surprisingly, we did not detect an overall trend in any of the calculated functional diversity metrics.
While nearly all of the studies included in our data experienced a significant increase or decrease in at least one of the metrics, trends were observed in equal numbers resulting in an overall trend that was not significantly different from zero.
As with previous species-based syntheses, we also found no overall trend in species diversity accompanied by increasing turnover through time [@dornelas2018], indicating that non-significant trends in functional metrics may be consistent with similar well-documented species derived trends.
We found no evidence of systemic functional richness loss or functional change.
A lack of trend for data broken down by realm, biome, and taxonomic group gives further evidence that directional functional change is absent from all systems observed in our dataset.
